\chapter{Introduction}
\label{chap:introduction}

A navigation-system, asked for the fastest route, usually returns a route over larger streets, like motorways.
When many drivers receive the same answer from their independent navigation-systems, the traffic tends to overload these larger streets.
These larger streets probably have multiple lanes and allow a higher speed-limit than smaller streets, resulting in a high driver-capacity and hence a high throughput.
In addition, due to rush-hours, the traffic is not balanced over the day.
In fact, a street-network can be quite loaded during rush-hour-scenarios.
Increasing the network's capacity often isn't sensible, because it reduces the overall efficiency in non-rush-hour-scenarios.
So, a typical street-network is a tradeoff between fitting the rush-hour-scenario and keeping the network efficient, when there is no rush-hour.
Hence these aspects lead to traffic-jams and former fast routes become inefficient due to their popularity.
Besides environmental impact and circumstances to handle the daily amount of rush-hour-traffic, traffic-jams in large (capital) cities cost days of the people's time and hence much money per year~(\cite{inrix:traffic-cost}).
Taking the popularity into account seems to be a mandatory step to reduce traffic-jams.
At the latest, this will be interesting with many self-driving cars, where many routes will be needed and driven automatically.
It would be inefficient to send all vehicles over the same streets.
So distributing this amount of traffic over the network is and will be important to save both, time and money.

This thesis explains, how a given street-network of multidimensional metrics (travel-distance, travel-time) can be augmented, such that diverse alternative paths can be found and used to spread the load (of a typical workload) while limiting the detours of individual \glspl{stpair}.
Although the presented methods can be transferred to different kinds of networks, this thesis focusses on street-networks.
A different kind of network can be wireless-networks or the Internet, where the focus lays more on nodes than edges.
Here, data-packages are transmitted much faster than drivers are driving through street-networks and traffic-jams occur in nodes, not in the edges.

\section{Related Work}

    In the following, dynamic and static methods are presented, that are currently used to solve these issues, having their pros and cons.
    This thesis picks up on some static methods.

    Dynamic (or adaptive) approaches describe the ability of vehicles finding new paths based on current conditions or circumstances, while driving along another path.
    So in opposite to static approaches, dynamic routing processes time-dependent, volatile data and processes it (more or less) immediately.
    Statically in this context means calculating the routes with help of non-volatile data.
    So, routing-computations are not dependent on a street-network's state, but on its attributes.
    Paths are calculated before a vehicle drives it, and the vehicles' path (typically) doesn't change over time.
    That's why a static approach is less dependent on dynamic traffic and more dependent on the street-network itself, in opposite to the dynamic approaches.

    \subsection{Dynamic routing}

        One example for a dynamic approach is Google, that has gathered~\cite{barth:google-traffic} anonymous data from Android-users to predict current traffic-jams in the street-network.
        Since most smartphones use Android~\cite{kantar:android-vs-ios}, there is enough data being collectable for sufficient accuracy.
        On the other hand, this approach requires a global observer, including infrastructure, being able to process incoming queries in reasonable time.
        With the drivers' data, like their location, the global observer can compute heuristics to approximate the current workload of other network-parts.
        Usually, this cost additional knowledge-exchange between the global observer (\eg\ Google) and the drivers.
        If the drivers aren't willing or able to share relevant information, the infrastructure has to be extended to collect relevant data.
        The need of extending the infrastructure makes the routing dependent on the infrastructure and could cost money.

    \subsection{Static routing}

        There are several static computation-methods to compute a selection of several routes with respect to only one metric, that are sufficiently distinct.
        One idea is the computation of the $k$ shortest paths for a given \gls{stpair}.
        According to~\cite{eppstein:finding_k_shortest_paths}, the computation is complex while resulting paths don't differ much.
        Another quite intuitive idea is the penalization of popular routes to reduce their usage before computing the shortest path again.
        The focus in \cite{bader:alternative-route-graphs} lays on measuring the quality of alternative paths via node-degrees and using heuristics for computing multidimensional shortest paths.
        In \cite{chondrogiannis:k_shortest_paths}, the problem is addressed, that paths of a set of found $k$ shortest paths for a given \gls{stpair} should differ from eachother, not only from the shortest path itself.
        Hence they build the set iteratively and discard unwished paths with a user-provided threshold.
        To this end, a similarity between paths is defined.
        However, both approaches suffer from the sensitivity of tuning the parameters and graphs are considered to have only one metric.

        For multiple metrics, another approach could be to distribute the found paths by using alternative routes.
        Here, the drivers would not necessarily get only the option of the fastest route, but also some other reasonable ones.
        When it comes to graphs of multidimensional metrics, pareto-optimal solutions occur implicitly.
        Thus, a first idea is computing all pareto-optimal paths.
        As stated in~\cite{delling:pareto-paths}, computing all pareto-optimal paths is too expensive to be practical at all.
        That's why~\cite{barth:alternative_multicriteria_routes} develops an algorithm, that enumerates paths with extending the search for shortest paths by a geometric interpretation in the metrics' underlying cost-space.
        To handle multidimensional metrics, the algorithm uses personalized routing (introduced in~\cite{geisberger:route_planning}), which takes predefined preferences to compute one scalar out of multiple metrics.
        To speed the queries up, a preprocessing-technique called contraction-hierarchies is applied in their paper.
        In addition, while the original approach of contraction-hierarchies is developed in~\cite{geisberger:contraction_hierarchies}, an adjusted version~\cite{funke:personal-routes} using linear programming is used for this thesis.
        The adjustment enables more efficient contraction-hierarchies when using personalized routing.
        Further, to reduce the search-space, the enumeration is restricted by a clever check from~\cite{barth:alternative_routes} without influencing its correctness or completeness.

\section{Main contribution}

    The contribution of this thesis is the combination of penalization and personalized routing.
    A computation-phase called balancing is added, that analyzes the street-network with the help of the \glsreset{repr}\gls{repr} from~\cite{barth:alternative_routes}.
    This preprocessing leads to a new metric, penalizing edges of high popularity.
    Applying personalized routing with all metrics, including the new one(s), leads to a more distributed set of routes.
    Thus, the street-network is used in a broader manner.
    At the same time, due to personalized routing, routing-queries can be processed as quickly as before.
    To guarantee a certain tolerance towards the shortest path, a filter is added to the \gls{repr}.

\section{Outline}

    This thesis is structured as following.
    \begin{description}
        \item[\cref{chap:preliminaries} -- \nameref{chap:preliminaries}:]
            This chapter explains the methods, that are combined in this thesis to create the new metric, that balances the provided graph to spread the found routes.
            These methods refer to a way of storing the graph in memory, describe the speedup of multicriteria-routing by contraction-hierarchies and a modified Dijkstra, and explain an algorithm to find alternative paths in a multicriteria setting.
        \item[\cref{chap:balancing} -- \nameref{chap:balancing}:]
            After the preliminaries are explained, the balancing is presented.
            The balancing is the core of this thesis and describes the looped chain of procedures, that creates the new metric.
        \item[\cref{chap:experiments} -- \nameref{chap:experiments}:]
            In this chapter, street-networsk from OpenStreetMap are used to show the balancing in action.
        \item[\cref{chap:conclusion} -- \nameref{chap:conclusion}:]
            Here, a summary is depicted and future work is discussed, such as performance-improvements and variations of balancing.
    \end{description}